---
title: "20- Raw WGBS data acquisition & QC"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

------------------------------------------------------------------------

# Overview

**First step of the self-contained `20+` pipeline.** This script downloads the
raw paired-end (2 x 150 bp) WGBS FastQs for all **24** *Mytilus trossulus*
mantle individuals from the `nightingales/M_trossulus` archive, verifies the
MD5 checksums, and produces a raw-read quality report with FastQC + MultiQC.

This step does not depend on any prior `output/` artifacts; it begins the
workflow from the original sequencing reads.

The 24 samples (mussel ID = `*M`) by PAH-exposure group:

| Group | Site codes | Samples |
|-------|------------|---------|
| Low PAH (treatment 0)  | HC, AP, BS | 105M 106M 107M 109M 239M 240M 241M 242M 269M 270M 271M 272M |
| High PAH (treatment 1) | EB, SA, SC | 69M 70M 71M 72M 78M 79M 80M 81M 92M 93M 94M 95M |

`93M` is retained here (24-sample set) for landscape/QC; it is dropped from the
23-sample differential set downstream (see `26-methylkit-object.Rmd`).

Output dir: `../output/20-raw-data-qc/`

```{bash make-output-dir}
mkdir -p ../output/20-raw-data-qc
mkdir -p ../data/raw-wgbs
```

# Download raw FastQs + checksums

Source: <https://owl.fish.washington.edu/nightingales/M_trossulus/>
(see `data/raw-wgbs/README.md`). Files are named `<number>M_[12].fastq.gz`.

```{bash download-raw-reads}
raw_reads_url="https://owl.fish.washington.edu/nightingales/M_trossulus/"
raw_reads_dir="../data/raw-wgbs"

wget \
--directory-prefix ${raw_reads_dir} \
--recursive \
--no-check-certificate \
--continue \
--no-directories \
--no-parent \
--quiet \
--level=1 \
--accept "[0-9]M*.fastq.gz,[0-9]M*.fastq.gz.md5sum" \
${raw_reads_url}

ls -lh ${raw_reads_dir}
```

# Verify checksums

```{bash verify-raw-read-checksums}
cd ../data/raw-wgbs

for file in *.md5sum
do
  md5sum --check "${file}"
done
```

# Confirm all 24 samples present

```{bash sample-inventory}
cd ../data/raw-wgbs
echo "R1 files:"; ls -1 *_1.fastq.gz | wc -l
echo "R2 files:"; ls -1 *_2.fastq.gz | wc -l
ls -1 *_1.fastq.gz | sed 's/_1.fastq.gz//' | sort
```

# FastQC + MultiQC on raw reads

```{bash raw-fastqc-multiqc}
raw_reads_dir="../data/raw-wgbs"
raw_fastqc_dir="../output/20-raw-data-qc"
threads=40

############ FASTQC ############
fastqc \
--threads ${threads} \
--outdir ${raw_fastqc_dir} \
--quiet \
${raw_reads_dir}/*.fastq.gz

############ MULTIQC ############
multiqc ${raw_fastqc_dir} -o ${raw_fastqc_dir} -n raw-multiqc

# Remove per-file FastQC zips to keep the output dir tidy
rm -f ${raw_fastqc_dir}/*_fastqc.zip

ls -lh ${raw_fastqc_dir}
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
