---
title: "21- fastp trimming & post-trim QC"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

------------------------------------------------------------------------

# Overview

Adapter/quality trimming of the 24 raw WGBS libraries downloaded in
`20-raw-data-qc.Rmd`, using [fastp](https://github.com/OpenGene/fastp).
Illumina adapters are auto-detected and the **first 20 bp of each read are hard
trimmed** from both R1 and R2 — these leading bases are the most biased in WGBS
libraries (composition + methylation-call artefacts). Trimmed reads are then
re-assessed with FastQC + MultiQC.

- Input: `../data/raw-wgbs/<sample>_[12].fastq.gz`
- Output: `../output/21-fastp-trimming/` (trimmed `*_R[12].fastp-trim.fq.gz`,
  fastp HTML/JSON reports, post-trim FastQC/MultiQC)

```{bash make-output-dir}
mkdir -p ../output/21-fastp-trimming/trimmed-fastqs
mkdir -p ../output/21-fastp-trimming/trimmed-fastqc
```

# fastp trimming (20 bp hard trim, both ends)

```{bash fastp-trimming}
raw_reads_dir="../data/raw-wgbs"
trimmed_dir="../output/21-fastp-trimming/trimmed-fastqs"
threads=40

cd ${raw_reads_dir}

for r1 in *_1.fastq.gz
do
  sample=$(basename "${r1}" _1.fastq.gz)
  r2="${sample}_2.fastq.gz"
  echo "Trimming ${sample}"

  fastp \
  --in1 "${r1}" \
  --in2 "${r2}" \
  --detect_adapter_for_pe \
  --trim_front1 20 \
  --trim_front2 20 \
  --thread ${threads} \
  --html "../../output/21-fastp-trimming/trimmed-fastqs/${sample}.fastp-trim.report.html" \
  --json "../../output/21-fastp-trimming/trimmed-fastqs/${sample}.fastp-trim.report.json" \
  --out1 "../../output/21-fastp-trimming/trimmed-fastqs/${sample}_R1.fastp-trim.fq.gz" \
  --out2 "../../output/21-fastp-trimming/trimmed-fastqs/${sample}_R2.fastp-trim.fq.gz" \
  2>> "../../output/21-fastp-trimming/trimmed-fastqs/fastp.stderr"
done

ls -lh ../../output/21-fastp-trimming/trimmed-fastqs/*.fq.gz
```

# Checksums for trimmed FastQs

```{bash trimmed-checksums}
cd ../output/21-fastp-trimming/trimmed-fastqs
for fq in *.fastp-trim.fq.gz
do
  md5sum "${fq}" > "${fq}.md5"
done
```

# FastQC + MultiQC on trimmed reads

```{bash trimmed-fastqc-multiqc}
trimmed_dir="../output/21-fastp-trimming/trimmed-fastqs"
trimmed_fastqc_dir="../output/21-fastp-trimming/trimmed-fastqc"
threads=40

############ FASTQC ############
fastqc \
--threads ${threads} \
--outdir ${trimmed_fastqc_dir} \
--quiet \
${trimmed_dir}/*.fastp-trim.fq.gz

############ MULTIQC ############
# Pull in fastp JSON reports too so MultiQC summarizes trimming
multiqc ${trimmed_fastqc_dir} ${trimmed_dir} -o ${trimmed_fastqc_dir} -n trimmed-multiqc

rm -f ${trimmed_fastqc_dir}/*_fastqc.zip
ls -lh ${trimmed_fastqc_dir}
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
