---
title: "22- Reference genome prep (Bismark)"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

------------------------------------------------------------------------

# Overview

Downloads the *M. trossulus* reference assembly **NCBI `GCF_036588685.1`**
(`PNRI_Mtr1.1.1.hap1`) plus its GFF annotation, then runs
`bismark_genome_preparation` to create the in-silico bisulfite-converted
Bowtie2 indices used for alignment in `23-bismark-align.Rmd`.

- Genome FASTA: `../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa`
- Annotation GFF: `../output/22-genome-prep/genomic.gff` (used downstream for
  feature distribution and DML/DMR annotation)
- Bismark index: `../output/22-genome-prep/Bisulfite_Genome/`

Tool paths follow the lab server convention used in `code/09-meth-quant.Rmd`
(`/home/shared/Bismark-0.24.0/`, `/home/shared/bowtie2-2.4.4-linux-x86_64`).

```{bash make-output-dir}
mkdir -p ../output/22-genome-prep
```

# Download genome + annotation (NCBI datasets)

```{bash download-genome}
cd ../output/22-genome-prep

# NCBI datasets CLI bundle (genome + GFF) for GCF_036588685.1
datasets download genome accession GCF_036588685.1 \
  --include genome,gff3 \
  --filename GCF_036588685.1.zip

unzip -o GCF_036588685.1.zip -d ncbi_dataset_dl

# Stage the FASTA and GFF at the top of the output dir under stable names
cp ncbi_dataset_dl/ncbi_dataset/data/GCF_036588685.1/*_genomic.fna \
   GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa
cp ncbi_dataset_dl/ncbi_dataset/data/GCF_036588685.1/genomic.gff \
   genomic.gff

ls -lh GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa genomic.gff
```

Alternative direct FTP fetch (if `datasets` CLI is unavailable):

```{bash download-genome-ftp}
cd ../output/22-genome-prep
base="https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/036/588/685/GCF_036588685.1_PNRI_Mtr1.1.1.hap1"

wget -q "${base}/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fna.gz"
wget -q "${base}/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.gff.gz"

gunzip -f GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fna.gz
gunzip -f GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.gff.gz

mv GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fna GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa
mv GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.gff genomic.gff
```

# Index FASTA + inspect annotation

```{bash faidx}
/home/shared/samtools-1.12/samtools faidx \
../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa

head ../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa.fai
```

```{bash gff-feature-types}
# Enumerate feature types present (used to build per-feature GFFs later)
awk '{if($3 && $3 !~ /^#/) print $3}' ../output/22-genome-prep/genomic.gff \
  | sort | uniq -c
```

# Bismark genome preparation

`bismark_genome_preparation` expects the genome FASTA to live alone in a
folder; it writes the C->T and G->A converted Bowtie2 indices into a
`Bisulfite_Genome/` subdirectory of that folder.

```{bash bismark-genome-prep}
/home/shared/Bismark-0.24.0/bismark_genome_preparation \
--path_to_aligner /home/shared/bowtie2-2.4.4-linux-x86_64 \
--verbose \
../output/22-genome-prep/

ls -R ../output/22-genome-prep/Bisulfite_Genome | head
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
