---
title: "23- Bismark alignment"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

------------------------------------------------------------------------

# Overview

Aligns the trimmed paired-end WGBS reads (`21-fastp-trimming.Rmd`) to the
bisulfite-converted `GCF_036588685.1` index built in `22-genome-prep.Rmd` with
Bismark + Bowtie2. A relaxed `--score_min L,0,-0.6` and `--non_directional`
mode are used (matching `code/01-bismark-init.sh` / the study pipeline). A
checkpoint log lets the loop resume after interruption.

- Trimmed reads: `../output/21-fastp-trimming/trimmed-fastqs/<sample>_R[12].fastp-trim.fq.gz`
- Genome folder (with `Bisulfite_Genome/`): `../output/22-genome-prep/`
- Output BAMs + reports: `../output/23-bismark-align/`

```{bash make-output-dir}
mkdir -p ../output/23-bismark-align
```

# Align all 24 samples

```{bash bismark-align}
reads_dir="../output/21-fastp-trimming/trimmed-fastqs"
genome_folder="../output/22-genome-prep/"
output_dir="../output/23-bismark-align"
checkpoint_file="../output/23-bismark-align/completed_samples.log"
score_min="L,0,-0.6"

touch ${checkpoint_file}

for r1 in ${reads_dir}/*_R1.fastp-trim.fq.gz; do
  sample_name=$(basename "${r1}" _R1.fastp-trim.fq.gz)

  # Skip if already completed
  if grep -q "^${sample_name}$" ${checkpoint_file}; then
    echo "Sample ${sample_name} already processed. Skipping..."
    continue
  fi

  echo "Aligning ${sample_name}"
  /home/shared/Bismark-0.24.0/bismark \
  --path_to_bowtie2 /home/shared/bowtie2-2.4.4-linux-x86_64 \
  -genome ${genome_folder} \
  -p 8 \
  -score_min ${score_min} \
  --non_directional \
  -1 ${reads_dir}/${sample_name}_R1.fastp-trim.fq.gz \
  -2 ${reads_dir}/${sample_name}_R2.fastp-trim.fq.gz \
  -o ${output_dir} \
  --basename ${sample_name} \
  > ${output_dir}/${sample_name}_stdout.log 2> ${output_dir}/${sample_name}_stderr.log

  if [ $? -eq 0 ]; then
    echo ${sample_name} >> ${checkpoint_file}
    echo "Sample ${sample_name} aligned."
  else
    echo "Sample ${sample_name} FAILED. See ${output_dir}/${sample_name}_stderr.log"
    break
  fi
done

ls -lh ${output_dir}/*_pe.bam
```

# Alignment summary report

```{bash bismark-summary}
cd ../output/23-bismark-align
/home/shared/Bismark-0.24.0/bismark2report
/home/shared/Bismark-0.24.0/bismark2summary

# Roll the per-sample Bismark reports into one MultiQC report
multiqc . -o . -n bismark-align-multiqc
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
