---
title: "24- Deduplication & BAM sort/index"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

------------------------------------------------------------------------

# Overview

Removes PCR duplicates from the Bismark paired-end BAMs (`23-bismark-align.Rmd`)
with `deduplicate_bismark`, then coordinate-sorts and indexes the deduplicated
BAMs with SAMtools. The sorted/indexed BAMs feed methylation extraction
(`25-meth-extract.Rmd`) and can also be loaded directly by methylKit
`processBismarkAln()` as an alternative entry point.

- Input: `../output/23-bismark-align/<sample>_pe.bam`
- Output: `../output/24-dedup-sort/<sample>_pe.deduplicated.bam`,
  `<sample>_pe.deduplicated.sorted.bam(.bai)`

```{bash make-output-dir}
mkdir -p ../output/24-dedup-sort
```

# Deduplicate (paired-end)

```{bash deduplicate}
find ../output/23-bismark-align/*_pe.bam \
| xargs -n 1 basename -s _pe.bam \
| parallel -j 8 /home/shared/Bismark-0.24.0/deduplicate_bismark \
--bam \
--paired \
--output_dir ../output/24-dedup-sort \
../output/23-bismark-align/{}_pe.bam

ls -lh ../output/24-dedup-sort/*deduplicated.bam
```

# Deduplication report (MultiQC)

```{bash dedup-report}
cd ../output/24-dedup-sort
/home/shared/Bismark-0.24.0/bismark2report
multiqc . -o . -n dedup-multiqc
```

# Sort + index deduplicated BAMs

```{bash sort-index}
cd ../output/24-dedup-sort

find *_pe.deduplicated.bam \
| xargs basename -s .bam \
| xargs -I{} /home/shared/samtools-1.12/samtools \
sort --threads 24 {}.bam \
-o {}.sorted.bam

# Index each sorted BAM
for bam in *_pe.deduplicated.sorted.bam
do
  /home/shared/samtools-1.12/samtools index "${bam}"
done

ls -lh *deduplicated.sorted.bam*
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
