---
title: "25- Methylation extraction & CpG coverage report"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

------------------------------------------------------------------------

# Overview

Extracts per-cytosine methylation calls from the deduplicated BAMs
(`24-dedup-sort.Rmd`) and generates the **merged-CpG coverage report** that is
the canonical input to methylKit for all downstream analysis.

`bismark_methylation_extractor` produces the bedGraph/`.cov`; `coverage2cytosine
--merge_CpG --zero_based` then combines the two strands of each CpG into a
single record, yielding `*.CpG_report.merged_CpG_evidence.cov` in Bismark `cov`
format (`chr  start  end  %meth  count_methylated  count_unmethylated`).

- Input: `../output/24-dedup-sort/<sample>_pe.deduplicated.bam`
- Genome folder: `../output/22-genome-prep/`
- Output: `../output/25-meth-extract/<sample>.CpG_report.merged_CpG_evidence.cov`

```{bash make-output-dir}
mkdir -p ../output/25-meth-extract
```

# Methylation extraction

```{bash methylation-extractor}
find ../output/24-dedup-sort/*_pe.deduplicated.bam \
| xargs -n 1 -I{} /home/shared/Bismark-0.24.0/bismark_methylation_extractor \
--bedGraph \
--counts \
--comprehensive \
--merge_non_CpG \
--multicore 24 \
--buffer_size 75% \
--output ../output/25-meth-extract \
{}

ls -lh ../output/25-meth-extract/*deduplicated.bismark.cov.gz
```

# coverage2cytosine — merged CpG evidence

```{bash coverage2cytosine}
find ../output/25-meth-extract/*deduplicated.bismark.cov.gz \
| xargs -n 1 basename -s _pe.deduplicated.bismark.cov.gz \
| parallel -j 24 /home/shared/Bismark-0.24.0/coverage2cytosine \
--genome_folder ../output/22-genome-prep/ \
-o ../output/25-meth-extract/{} \
--merge_CpG \
--zero_based \
../output/25-meth-extract/{}_pe.deduplicated.bismark.cov.gz

ls -lh ../output/25-meth-extract/*.CpG_report.merged_CpG_evidence.cov
```

# Sanity check cov format + confirm 24 samples

```{bash inspect-cov}
echo "Merged-CpG cov files:"
ls -1 ../output/25-meth-extract/*.CpG_report.merged_CpG_evidence.cov | wc -l
ls -1 ../output/25-meth-extract/*.CpG_report.merged_CpG_evidence.cov \
  | xargs -n1 basename | sed 's/.CpG_report.merged_CpG_evidence.cov//' | sort

echo ""
echo "Example record (chr start end %meth count_m count_u):"
head -n 3 ../output/25-meth-extract/105M.CpG_report.merged_CpG_evidence.cov
```

# Extraction QC report

```{bash extraction-multiqc}
cd ../output/25-meth-extract
/home/shared/Bismark-0.24.0/bismark2report
multiqc . -o . -n meth-extract-multiqc
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
