---
title: "25.1- Recover 93M merged-CpG coverage file (full pipeline)"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

------------------------------------------------------------------------

# Overview

Sample **93M** is the only individual missing a
`*.CpG_report.merged_CpG_evidence.cov` file. Investigation showed **93M was
never fully aligned**:

- The gannet `data/` mirror holds 93M's trimmed FASTQs (`93M_R1/R2.fastp-trim.fq.gz`,
  ~4.5 GB each — comparable to every other sample) but **no BAM and no cov**.
- The repo copy `../output/04-bismark-pipeline/93M_*` is a **10,000-read test
  run** (`Sequence pairs analysed in total: 10000`, from the `-u 10000` bismark
  run noted in `05-bismark-cov.Rmd`). Merging that subset gives only ~17.7k CpGs
  — a 761 KB file versus the real ~795 MB files (69M/92M). **Do not use it.**

So a *complete* 93M cov requires running the production pipeline (scripts
`22`→`25`) for this one sample, with the exact parameters used for the other 23
so 93M is directly comparable. The executable driver is
[`25.1-93M-cov-recovery.sh`](25.1-93M-cov-recovery.sh); the chunks below mirror
it. Run it in the background:

```{bash run-driver}
nohup bash 25.1-93M-cov-recovery.sh \
  > ../output/25.1-93M-cov-recovery/run.log 2>&1 &
```

Pipeline, all matching scripts `22`–`25`:

| Step | Tool | Key params |
|------|------|-----------|
| index | `bismark_genome_preparation` | `GCF_036588685.1` |
| align | `bismark` (Bowtie2) | `-score_min L,0,-0.6 --non_directional`, paired |
| dedup | `deduplicate_bismark` | `--paired` |
| extract | `bismark_methylation_extractor` | `--bedGraph --comprehensive --merge_non_CpG` |
| merge | `coverage2cytosine` | `--merge_CpG --zero_based` |

- Reads:  `../output/25.1-93M-cov-recovery/reads/93M_R[12].fastp-trim.fq.gz` (from gannet)
- Genome: `../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa` (NCBI; gannet `.fa` returns 403)
- Output: `../output/25-meth-extract/93M.CpG_report.merged_CpG_evidence.cov`

```{bash make-dirs}
mkdir -p ../output/22-genome-prep ../output/23-bismark-align \
         ../output/24-dedup-sort ../output/25-meth-extract \
         ../output/25.1-93M-cov-recovery/reads
```

# 1. Trimmed reads

```{bash fetch-reads}
g="https://gannet.fish.washington.edu/seashell/bu-github/project-mytilus-methylation/data"
d=../output/25.1-93M-cov-recovery/reads
for p in R1 R2; do
  [ -s "$d/93M_${p}.fastp-trim.fq.gz" ] || \
    wget -q -O "$d/93M_${p}.fastp-trim.fq.gz" "$g/93M_${p}.fastp-trim.fq.gz"
  gzip -t "$d/93M_${p}.fastp-trim.fq.gz"
done
```

# 2. Reference genome + bisulfite index

The FASTA is fetched from NCBI (assembly `GCF_036588685.1`; RefSeq `NC_…`
accessions match the `.fai`). `bismark_genome_preparation` builds the
`Bisulfite_Genome/` Bowtie2 indices.

```{bash genome-prep}
fa=../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa
ncbi="https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/036/588/685/GCF_036588685.1_PNRI_Mtr1.1.1.hap1/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fna.gz"
if [ ! -s "$fa" ]; then wget -q -O "$fa.gz" "$ncbi"; gunzip -f "$fa.gz"; fi

[ -d ../output/22-genome-prep/Bisulfite_Genome ] || \
/home/shared/Bismark-0.24.0/bismark_genome_preparation \
  --path_to_aligner /home/shared/bowtie2-2.4.4-linux-x86_64 \
  --verbose ../output/22-genome-prep/
```

# 3. Bismark alignment (paired, non-directional)

```{bash align}
# Bismark 0.24 forbids --basename with --multicore; -p 12 saturates the box
# because non-directional mode already runs 4 concurrent bowtie2 instances.
/home/shared/Bismark-0.24.0/bismark \
  --path_to_bowtie2 /home/shared/bowtie2-2.4.4-linux-x86_64 \
  -genome ../output/22-genome-prep/ \
  -p 12 \
  -score_min L,0,-0.6 \
  --non_directional \
  -1 ../output/25.1-93M-cov-recovery/reads/93M_R1.fastp-trim.fq.gz \
  -2 ../output/25.1-93M-cov-recovery/reads/93M_R2.fastp-trim.fq.gz \
  -o ../output/23-bismark-align \
  --basename 93M
```

# 4. Deduplicate

```{bash dedup}
/home/shared/Bismark-0.24.0/deduplicate_bismark --bam --paired \
  --output_dir ../output/24-dedup-sort \
  ../output/23-bismark-align/93M_pe.bam
```

# 5. Methylation extraction

```{bash extract}
/home/shared/Bismark-0.24.0/bismark_methylation_extractor \
  --bedGraph --counts --comprehensive --merge_non_CpG \
  --multicore 24 --buffer_size 50% \
  --output ../output/25-meth-extract \
  ../output/24-dedup-sort/93M_pe.deduplicated.bam
```

# 6. coverage2cytosine — merged CpG evidence

```{bash merge-cpg}
/home/shared/Bismark-0.24.0/coverage2cytosine \
  --genome_folder ../output/22-genome-prep/ \
  -o ../output/25-meth-extract/93M \
  --merge_CpG --zero_based \
  ../output/25-meth-extract/93M_pe.deduplicated.bismark.cov.gz

rm -f ../output/25-meth-extract/93M.CpG_report.txt   # drop multi-GB intermediate
```

# 7. Sanity check

Real samples produce ~30–40M CpGs (~795 MB). Confirm 93M is now on that scale,
in 6-column Bismark cov format, before feeding it to `26-methylkit-object.Rmd`.

```{bash inspect}
cov=../output/25-meth-extract/93M.CpG_report.merged_CpG_evidence.cov
ls -lh "$cov"
echo "records: $(wc -l < "$cov")"
head -n 3 "$cov"
awk '{m+=$5;u+=$6} END{printf "global weighted %%CpG meth: %.2f%%  (%d CpG)\n",100*m/(m+u),NR}' "$cov"
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
