---
title: "26- methylKit object construction & filtering"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(methylKit)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

------------------------------------------------------------------------

# Overview

Builds the shared methylKit objects consumed by every downstream analysis from
the merged-CpG `cov` files produced in `25-meth-extract.Rmd`. Two sample sets
are constructed and saved as `.rds` so later steps are resumable
(per the plan's reproducibility notes):

| Object | Set | Samples | Use |
|--------|-----|---------|-----|
| `myobj_24` | 24-sample | all incl. **93M** | Goal 1 landscape / QC (`27`–`29`) |
| `myobj_23` | 23-sample | 24 minus **93M** | Goal 2 differential / biomarkers (`30`–`33`) |

Treatment coding (per `plan.md`, `code/10`/`code/11`): low-PAH **HC, AP, BS = 0**;
high-PAH **EB, SA, SC = 1**.

Coverage filtering follows the plan: **10x primary** (`lo.count = 10`) with a
**5x sensitivity** pass, and `hi.perc = 98` to cap PCR/duplication artefacts.

- Input: `../output/25-meth-extract/<sample>.CpG_report.merged_CpG_evidence.cov`
- Output: `../output/26-methylkit-object/` (`.rds` objects, filtering tables)

```{bash make-output-dir}
mkdir -p ../output/26-methylkit-object ../output/25-meth-extract
```

# Fetch the per-sample cov files

The 23 published merged-CpG cov files live on gannet; **93M** is generated
locally in `25.1-93M-cov-recovery` and already sits in `../output/25-meth-extract`.
Pull anything not already present (~770 MB each, idempotent).

```{bash fetch-cov}
cov_dir="../output/25-meth-extract"
gannet="https://gannet.fish.washington.edu/seashell/bu-github/project-mytilus-methylation/data"

# every sample except 93M (which is built locally)
for s in 105M 106M 107M 109M 239M 240M 241M 242M 269M 270M 271M 272M \
         69M 70M 71M 72M 78M 79M 80M 81M 92M 94M 95M; do
  f="$cov_dir/${s}.CpG_report.merged_CpG_evidence.cov"
  [ -s "$f" ] || { echo "downloading ${s}"; wget -q -O "$f" "$gannet/${s}.CpG_report.merged_CpG_evidence.cov"; }
done

echo "cov files present: $(ls -1 "$cov_dir"/*.CpG_report.merged_CpG_evidence.cov | wc -l) (expect 24)"
```

```{r install-methylkit}
# methylKit is expected to be preinstalled; install only if missing.
if (!requireNamespace("methylKit", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  BiocManager::install("methylKit")
}
```

# Sample sheet

```{r sample-sheet}
cov_dir <- "../output/25-meth-extract"

# 24-sample set (landscape). Order groups low (0) then high (1); 93M retained.
samples_24 <- c("105M","106M","107M","109M",   # HC  (low, 0)
                "239M","240M","241M","242M",   # AP  (low, 0)
                "269M","270M","271M","272M",   # BS  (low, 0)
                "69M","70M","71M","72M",       # EB  (high, 1)
                "78M","79M","80M","81M",       # SA  (high, 1)
                "92M","93M","94M","95M")       # SC  (high, 1)

site_24 <- c(rep("HC",4), rep("AP",4), rep("BS",4),
             rep("EB",4), rep("SA",4), rep("SC",4))
pah_24  <- ifelse(site_24 %in% c("HC","AP","BS"), 0L, 1L)

# 23-sample set (differential): drop 93M (coverage/quality outlier)
keep_23   <- samples_24 != "93M"
samples_23 <- samples_24[keep_23]
site_23    <- site_24[keep_23]
pah_23     <- pah_24[keep_23]

meta <- data.frame(sample = samples_24, site = site_24, pah = pah_24)
write.csv(meta, "../output/26-methylkit-object/sample_metadata.csv", row.names = FALSE)
meta
```

```{r cov-file-lists}
cov_path <- function(s) file.path(cov_dir, paste0(s, ".CpG_report.merged_CpG_evidence.cov"))

files_24 <- as.list(cov_path(samples_24))
files_23 <- as.list(cov_path(samples_23))

stopifnot(all(file.exists(unlist(files_24))))
```

# Read 24-sample set (landscape, descriptive mincov = 1)

For the descriptive landscape pass we read with `mincov = 1` so low-coverage
sites remain visible before any filtering (per plan 1.1).

```{r read-24}
myobj_24 <- methRead(
  location   = files_24,
  sample.id  = as.list(samples_24),
  assembly   = "GCF_036588685.1",
  pipeline   = "bismarkCoverage",
  context    = "CpG",
  mincov     = 1,
  treatment  = pah_24
)

saveRDS(myobj_24, "../output/26-methylkit-object/myobj_24.rds")
```

# Read 23-sample set (differential, mincov = 2)

```{r read-23}
myobj_23 <- methRead(
  location   = files_23,
  sample.id  = as.list(samples_23),
  assembly   = "GCF_036588685.1",
  pipeline   = "bismarkCoverage",
  context    = "CpG",
  mincov     = 2,
  treatment  = pah_23
)

saveRDS(myobj_23, "../output/26-methylkit-object/myobj_23.rds")
```

# Coverage filtering — 10x primary + 5x sensitivity

```{r filter-23}
myobj_23 <- readRDS("../output/26-methylkit-object/myobj_23.rds")

# Primary: 10x minimum, cap top 2% of coverage
filt_23_10x <- filterByCoverage(myobj_23,
                                lo.count = 10, lo.perc = NULL,
                                hi.count = NULL, hi.perc = 98)

# Sensitivity: 5x minimum
filt_23_5x  <- filterByCoverage(myobj_23,
                                lo.count = 5,  lo.perc = NULL,
                                hi.count = NULL, hi.perc = 98)

saveRDS(filt_23_10x, "../output/26-methylkit-object/filt_23_10x.rds")
saveRDS(filt_23_5x,  "../output/26-methylkit-object/filt_23_5x.rds")
```

# unite (require coverage in >=5 individuals/group)

cov files are already CpG-merged, so `destrand = FALSE` is correct (destranding
only applies to per-strand cytosine reports).

```{r unite-23}
meth_23_10x <- unite(filt_23_10x, min.per.group = 5L, destrand = FALSE)
meth_23_5x  <- unite(filt_23_5x,  min.per.group = 5L, destrand = FALSE)

saveRDS(meth_23_10x, "../output/26-methylkit-object/meth_23_10x.rds")
saveRDS(meth_23_5x,  "../output/26-methylkit-object/meth_23_5x.rds")

c(sites_10x = nrow(meth_23_10x), sites_5x = nrow(meth_23_5x))
```

# CpG-retention trade-off table (1x / 5x / 10x)

Transparent reporting of how many CpGs survive each threshold (plan 1.5 / 2.x).

```{r retention-table}
retention <- function(obj, thresholds = c(1, 5, 10)) {
  do.call(rbind, lapply(seq_along(obj), function(i) {
    cov <- getData(obj[[i]])$coverage
    data.frame(sample = obj[[i]]@sample.id,
               t(sapply(thresholds, function(t) sum(cov >= t))))
  })) |>
    setNames(c("sample", paste0("CpG_ge_", thresholds, "x")))
}

ret <- retention(myobj_23)
write.csv(ret, "../output/26-methylkit-object/cpg_retention_by_threshold.csv", row.names = FALSE)
ret
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
