---
title: "27- Methylation landscape (Goal 1)"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(methylKit)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

------------------------------------------------------------------------

# Overview — Goal 1

Genome-wide, treatment-independent characterization of the *M. trossulus* CpG
methylome plus the sample-level QC baseline, using the **24-sample** object
(`myobj_24`, incl. 93M) from `26-methylkit-object.Rmd`. Implements plan
sections 1.1, 1.2 and 1.5 (feature distribution is `28`, CpG O/E is `29`).

Output dir: `../output/27-methylation-landscape/`

```{bash make-output-dir}
mkdir -p ../output/27-methylation-landscape
```

```{r load-object}
myobj_24 <- readRDS("../output/26-methylkit-object/myobj_24.rds")
meta     <- read.csv("../output/26-methylkit-object/sample_metadata.csv")
```

# 1.1 Per-sample global CpG methylation summary

Global **weighted** % CpG methylation = Σ methylated counts / Σ total counts.

```{r per-sample-global}
per_sample <- do.call(rbind, lapply(seq_along(myobj_24), function(i) {
  d  <- getData(myobj_24[[i]])
  tot <- d$coverage
  meth <- d$numCs
  data.frame(
    sample          = myobj_24[[i]]@sample.id,
    n_CpG_covered   = nrow(d),
    mean_coverage   = mean(tot),
    median_coverage = median(tot),
    global_pct_meth = 100 * sum(meth) / sum(tot)
  )
})) |>
  left_join(meta, by = "sample")

write.csv(per_sample, "../output/27-methylation-landscape/per_sample_global_methylation.csv",
          row.names = FALSE)
per_sample
```

```{r global-mean-check}
# Manuscript reports global mean ~12% (range 10.3-13%); confirm here.
summary(per_sample$global_pct_meth)
```

# Per-sample methylation & coverage histograms

```{r stats-plots}
pdf("../output/27-methylation-landscape/methylation_stats_histograms.pdf", width = 6, height = 4)
for (i in seq_along(myobj_24)) {
  getMethylationStats(myobj_24[[i]], plot = TRUE, both.strands = FALSE)
}
dev.off()

pdf("../output/27-methylation-landscape/coverage_stats_histograms.pdf", width = 6, height = 4)
for (i in seq_along(myobj_24)) {
  getCoverageStats(myobj_24[[i]], plot = TRUE, both.strands = FALSE)
}
dev.off()
```

# 1.2 Methylation-level distribution & CpG classification

Bivalve methylomes are bimodal/mosaic. Classify sufficiently covered CpGs
(>=10x) into three standard invertebrate bins.

```{r classify-cpgs}
classify <- function(obj, mincov = 10) {
  do.call(rbind, lapply(seq_along(obj), function(i) {
    d <- getData(obj[[i]])
    d <- d[d$coverage >= mincov, ]
    pm <- 100 * d$numCs / d$coverage
    cls <- cut(pm, breaks = c(-Inf, 10, 50, Inf),
               labels = c("unmethylated", "intermediate", "methylated"))
    as.data.frame(table(cls)) |>
      mutate(sample = obj[[i]]@sample.id,
             prop   = Freq / sum(Freq))
  }))
}

cpg_classes <- classify(myobj_24, mincov = 10)
write.csv(cpg_classes, "../output/27-methylation-landscape/cpg_class_proportions.csv",
          row.names = FALSE)

# Genome-wide pooled proportions
cpg_classes |> group_by(cls) |> summarise(prop = sum(Freq) / sum(cpg_classes$Freq))
```

```{r bimodality-plot}
# Pooled per-CpG % methylation distribution (>=10x), showing bimodality
pm_all <- unlist(lapply(seq_along(myobj_24), function(i) {
  d <- getData(myobj_24[[i]]); d <- d[d$coverage >= 10, ]
  100 * d$numCs / d$coverage
}))

p <- ggplot(data.frame(pct = pm_all), aes(pct)) +
  geom_histogram(bins = 50, fill = "grey30") +
  labs(x = "% CpG methylation (>=10x, pooled)", y = "CpG count",
       title = "Genome-wide methylation bimodality") +
  theme_minimal()
ggsave("../output/27-methylation-landscape/bimodality_pooled.png", p, width = 6, height = 4)
```

# 1.5 Sample-level QC: correlation, clustering, PCA

Use the united matrix at 10x (from `26-methylkit-object.Rmd`, 23-sample) plus a
24-sample unite to position **93M**.

```{r unite-24-for-qc}
# Light filter on the 24-sample set for global structure (10x, cap 98%)
filt_24 <- filterByCoverage(myobj_24, lo.count = 10, lo.perc = NULL,
                            hi.count = NULL, hi.perc = 98)
meth_24 <- unite(filt_24, min.per.group = 4L, destrand = FALSE)
saveRDS(meth_24, "../output/27-methylation-landscape/meth_24_10x.rds")
nrow(meth_24)
```

```{r correlation}
png("../output/27-methylation-landscape/sample_correlation.png", width = 1200, height = 1200)
getCorrelation(meth_24, plot = TRUE)
dev.off()
```

```{r clustering}
png("../output/27-methylation-landscape/sample_clustering.png", width = 1000, height = 700)
clusterSamples(meth_24, dist = "correlation", method = "ward", plot = TRUE)
dev.off()
```

```{r pca}
png("../output/27-methylation-landscape/sample_pca.png", width = 800, height = 700)
PCASamples(meth_24)
dev.off()

# PCA scores for inspecting whether samples group by site vs PAH vs artefact
pca_obj <- PCASamples(meth_24, obj.return = TRUE)
write.csv(as.data.frame(pca_obj$x),
          "../output/27-methylation-landscape/pca_scores.csv")
```

# 93M outlier evaluation

Formal evaluation of 93M against the QC criteria: global %CpG methylation,
coverage, and PCA/clustering position. Document the retain/exclude decision
with the figures above rather than as an unstated convention.

```{r evaluate-93M}
per_sample |>
  mutate(is_93M = sample == "93M") |>
  arrange(global_pct_meth) |>
  select(sample, site, pah, n_CpG_covered, mean_coverage, global_pct_meth, is_93M)
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
