---
title: "28- Genomic-feature distribution (Goal 1)"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(methylKit)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

------------------------------------------------------------------------

# Overview — Goal 1.3

Distributes covered CpGs (and their methylation levels) across genomic features
to demonstrate the canonical molluscan **gene-body–enriched mosaic** pattern.
Builds feature BEDs from the `GCF_036588685.1` GFF (`22-genome-prep.Rmd`) with
BEDTools, intersects covered CpG positions against each feature class, and
summarizes CpG fraction + mean methylation per feature.

- GFF: `../output/22-genome-prep/genomic.gff`
- Genome index: `../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa.fai`
- 24-sample object: `../output/26-methylkit-object/myobj_24.rds`
- Output dir: `../output/28-genomic-feature-distribution/`

```{bash make-output-dir}
mkdir -p ../output/28-genomic-feature-distribution/features
```

# Annotation GFF

Fetch the `GCF_036588685.1` GFF from NCBI if it is not already staged in
`22-genome-prep` (the gannet genome mirror serves the FASTA 403; NCBI is used
for both). Idempotent.

```{bash fetch-gff}
gff="../output/22-genome-prep/genomic.gff"
if [ ! -s "$gff" ]; then
  url="https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/036/588/685/GCF_036588685.1_PNRI_Mtr1.1.1.hap1/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.gff.gz"
  wget -q -O "$gff.gz" "$url" && gunzip -f "$gff.gz"
fi
ls -lh "$gff"
```

# Per-feature GFFs from the annotation

```{bash split-gff}
input_gff="../output/22-genome-prep/genomic.gff"
out_dir="../output/28-genomic-feature-distribution/features"

features=("gene" "mRNA" "exon" "CDS" "lnc_RNA" "ncRNA" "tRNA" "rRNA" "pseudogene")

for feature in "${features[@]}"; do
  awk -v ft="$feature" '$3 == ft && $1 !~ /^#/ {print}' "$input_gff" \
    > "$out_dir/${feature}.gff"
  echo "Created $out_dir/${feature}.gff ($(wc -l < "$out_dir/${feature}.gff") records)"
done
```

# Derived contexts (intron / promoter / intergenic)

```{bash genome-file}
# BEDTools genome file from the FASTA index (chrom <tab> length).
# Index the FASTA first if it was not indexed in 22-genome-prep.
fa=../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa
[ -s "$fa.fai" ] || /home/shared/samtools-1.12/samtools faidx "$fa"

cut -f1,2 "$fa.fai" \
  | sort -k1,1 > ../output/28-genomic-feature-distribution/genome.txt
wc -l ../output/28-genomic-feature-distribution/genome.txt
head ../output/28-genomic-feature-distribution/genome.txt
```

```{bash derived-features}
feat="../output/28-genomic-feature-distribution/features"
genome="../output/28-genomic-feature-distribution/genome.txt"

# GFF -> sorted BED helper (cols 1,4-1,5 -> 0-based start)
gff2bed () { awk 'BEGIN{OFS="\t"} $1!~/^#/ {print $1, $4-1, $5}' "$1" | sort -k1,1 -k2,2n; }

gff2bed "$feat/gene.gff" > "$feat/gene.bed"
gff2bed "$feat/exon.gff" > "$feat/exon.bed"
gff2bed "$feat/mRNA.gff" > "$feat/mRNA.bed"

# Introns = gene minus exon
bedtools subtract -a "$feat/gene.bed" -b "$feat/exon.bed" > "$feat/intron.bed"

# Promoters/TSS = 1 kb upstream flank of genes (strand-aware via gene GFF)
awk 'BEGIN{OFS="\t"} $1!~/^#/ {print $1,$4-1,$5,".",".",$7}' "$feat/gene.gff" \
  | sort -k1,1 -k2,2n > "$feat/gene_stranded.bed"
bedtools flank -i "$feat/gene_stranded.bed" -g "$genome" -l 1000 -r 0 -s \
  | cut -f1-3 | sort -k1,1 -k2,2n > "$feat/promoter_1kb.bed"

# Intergenic = genome complement of genes
bedtools complement -i "$feat/gene.bed" -g "$genome" > "$feat/intergenic.bed"

wc -l "$feat"/{gene,exon,intron,promoter_1kb,intergenic}.bed
```

> Repeats: if a RepeatMasker `.out`/`.bed` track becomes available, add a
> `repeat.bed` here and include it below; otherwise note as an annotation gap.

# Covered CpG positions -> BED (>=10x, pooled across 24 samples)

```{r covered-cpg-bed}
myobj_24 <- readRDS("../output/26-methylkit-object/myobj_24.rds")

# Pool: a CpG is "covered" if >=10x in any sample; record mean % meth across
# samples meeting that threshold.
cpg_tab <- lapply(seq_along(myobj_24), function(i) {
  d <- getData(myobj_24[[i]]); d <- d[d$coverage >= 10, ]
  data.frame(chr = d$chr, start = d$start, end = d$end,
             pct = 100 * d$numCs / d$coverage)
}) |> bind_rows() |>
  group_by(chr, start, end) |>
  summarise(pct_meth = mean(pct), .groups = "drop")

options(scipen = 999)   # never write BED coordinates in scientific notation
bed <- cpg_tab |>
  transmute(chr,
            start = as.integer(start - 1),   # integers can't render as 1.06e+08
            end   = as.integer(end),
            pct_meth) |>
  arrange(chr, start)

write.table(bed, "../output/28-genomic-feature-distribution/covered_cpg_10x.bed",
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
nrow(bed)
```

# Intersect covered CpGs against each feature

```{bash intersect-features}
feat="../output/28-genomic-feature-distribution/features"
cpg="../output/28-genomic-feature-distribution/covered_cpg_10x.bed"
out="../output/28-genomic-feature-distribution"

sort -k1,1 -k2,2n "$cpg" > "$out/covered_cpg_10x.sorted.bed"

for f in gene exon intron mRNA promoter_1kb intergenic; do
  bedtools intersect -a "$out/covered_cpg_10x.sorted.bed" -b "$feat/${f}.bed" -u \
    > "$out/cpg_in_${f}.bed"
  echo "${f} $(wc -l < "$out/cpg_in_${f}.bed")"
done
```

# Feature x methylation-level summary + barplot

```{r feature-summary}
out <- "../output/28-genomic-feature-distribution"
features <- c("gene","exon","intron","mRNA","promoter_1kb","intergenic")

total_cpg <- nrow(read.table(file.path(out, "covered_cpg_10x.sorted.bed")))

feat_summary <- map_dfr(features, function(f) {
  fp <- file.path(out, paste0("cpg_in_", f, ".bed"))
  if (file.size(fp) == 0) return(tibble(feature = f, n_cpg = 0,
                                         frac_cpg = 0, mean_pct_meth = NA))
  d <- read.table(fp)
  tibble(feature = f, n_cpg = nrow(d),
         frac_cpg = nrow(d) / total_cpg,
         mean_pct_meth = mean(d$V4))
})

write.csv(feat_summary, file.path(out, "feature_methylation_summary.csv"),
          row.names = FALSE)

p <- ggplot(feat_summary, aes(reorder(feature, -mean_pct_meth), mean_pct_meth)) +
  geom_col(fill = "steelblue") +
  labs(x = "feature", y = "mean % CpG methylation",
       title = "Methylation by genomic feature (expect gene-body enrichment)") +
  theme_minimal()
ggsave(file.path(out, "feature_methylation_barplot.png"), p, width = 6, height = 4)
feat_summary
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
