---
title: "29- CpG O/E mosaic structure (Goal 1)"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(Biostrings)
library(GenomicRanges)
library(rtracklayer)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

------------------------------------------------------------------------

# Overview — Goal 1.4

Computes **CpG observed/expected (CpG O/E)** per gene/CDS from the reference
FASTA to characterize the mosaic genome architecture, then relates CpG O/E to
observed methylation. Bivalve gene sets typically show a **bimodal** CpG O/E
distribution; the low-O/E mode corresponds to historically germline-methylated,
constitutively expressed, more heavily methylated gene bodies.

\[ \text{CpG O/E} = \frac{N_{CpG}}{N_C \times N_G} \times \frac{L^2}{L-1} \]

- Genome FASTA: `../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa`
- Gene GFF: `../output/22-genome-prep/genomic.gff`
- Observed methylation: `../output/26-methylkit-object/myobj_24.rds`
- Output dir: `../output/29-cpg-oe-mosaic/`

```{bash make-output-dir}
mkdir -p ../output/29-cpg-oe-mosaic
```

# Load genome + gene coordinates

```{r load-genome}
genome <- readDNAStringSet("../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa")
# Normalize names to first token (matches GFF seqid)
names(genome) <- sub("\\s.*$", "", names(genome))

gff   <- import("../output/22-genome-prep/genomic.gff")
genes <- gff[gff$type == "gene"]
genes <- genes[as.character(seqnames(genes)) %in% names(genome)]
length(genes)
```

# CpG O/E per gene

```{r cpg-oe}
cpg_oe <- function(seq) {
  s  <- as.character(seq)
  L  <- nchar(s)
  if (L < 2) return(NA_real_)
  nC   <- lengths(gregexpr("C", s, fixed = TRUE)); nC <- ifelse(grepl("C", s), nC, 0)
  nG   <- lengths(gregexpr("G", s, fixed = TRUE)); nG <- ifelse(grepl("G", s), nG, 0)
  nCpG <- lengths(gregexpr("CG", s, fixed = TRUE)); nCpG <- ifelse(grepl("CG", s), nCpG, 0)
  if (nC == 0 || nG == 0) return(NA_real_)
  (nCpG / (nC * nG)) * (L^2 / (L - 1))
}

gene_seqs <- getSeq(genome, genes)
oe <- vapply(seq_along(gene_seqs), function(i) cpg_oe(gene_seqs[[i]]), numeric(1))

gene_oe <- tibble(
  gene_id = genes$ID,
  chr     = as.character(seqnames(genes)),
  start   = start(genes),
  end     = end(genes),
  width   = width(genes),
  cpg_oe  = oe
) |> filter(!is.na(cpg_oe))

write.csv(gene_oe, "../output/29-cpg-oe-mosaic/gene_cpg_oe.csv", row.names = FALSE)
summary(gene_oe$cpg_oe)
```

# Bimodal CpG O/E distribution

```{r oe-distribution}
p <- ggplot(gene_oe, aes(cpg_oe)) +
  geom_histogram(bins = 60, fill = "grey30") +
  geom_vline(xintercept = median(gene_oe$cpg_oe), linetype = 2, colour = "red") +
  labs(x = "CpG O/E (per gene)", y = "gene count",
       title = "Per-gene CpG O/E distribution (expect bimodality)") +
  theme_minimal()
ggsave("../output/29-cpg-oe-mosaic/cpg_oe_distribution.png", p, width = 6, height = 4)
```

```{r oe-modes}
# Optional 2-component mixture to formalize the two modes
if (requireNamespace("mixtools", quietly = TRUE)) {
  set.seed(1)
  mix <- mixtools::normalmixEM(gene_oe$cpg_oe, k = 2)
  saveRDS(mix, "../output/29-cpg-oe-mosaic/cpg_oe_mixture.rds")
  data.frame(mode = 1:2, mean = mix$mu, sd = mix$sigma, lambda = mix$lambda)
}
```

# Relate CpG O/E to observed methylation

Aggregate observed per-gene methylation (mean over covered CpGs, 24-sample
pooled >=10x) and join to CpG O/E. Expect low-O/E genes to be more methylated.

```{r oe-vs-methylation}
library(methylKit)
myobj_24 <- readRDS("../output/26-methylkit-object/myobj_24.rds")

cpg_meth <- lapply(seq_along(myobj_24), function(i) {
  d <- getData(myobj_24[[i]]); d <- d[d$coverage >= 10, ]
  data.frame(chr = d$chr, start = d$start, pct = 100 * d$numCs / d$coverage)
}) |> bind_rows()

cpg_gr  <- GRanges(cpg_meth$chr, IRanges(cpg_meth$start, width = 1))
gene_gr <- GRanges(gene_oe$chr, IRanges(gene_oe$start, gene_oe$end), gene_id = gene_oe$gene_id)

hits <- findOverlaps(cpg_gr, gene_gr)
gene_meth <- tibble(gene_id = gene_gr$gene_id[subjectHits(hits)],
                    pct     = cpg_meth$pct[queryHits(hits)]) |>
  group_by(gene_id) |>
  summarise(gene_mean_meth = mean(pct), n_cpg = n(), .groups = "drop")

oe_meth <- inner_join(gene_oe, gene_meth, by = "gene_id")
write.csv(oe_meth, "../output/29-cpg-oe-mosaic/cpg_oe_vs_methylation.csv", row.names = FALSE)

p <- ggplot(oe_meth, aes(cpg_oe, gene_mean_meth)) +
  geom_point(alpha = 0.2, size = 0.6) +
  geom_smooth(method = "loess") +
  labs(x = "CpG O/E", y = "mean gene-body % methylation",
       title = "Low-O/E genes are the methylation-prone gene bodies") +
  theme_minimal()
ggsave("../output/29-cpg-oe-mosaic/oe_vs_methylation.png", p, width = 6, height = 4)

cor.test(oe_meth$cpg_oe, oe_meth$gene_mean_meth, method = "spearman")
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
