---
title: "30- DML & DMR (Goal 2)"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(methylKit)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

------------------------------------------------------------------------

# Overview — Goal 2 (2.1-2.3, 2.5)

Differential methylation between **low-PAH (0)** and **high-PAH (1)** mussels on
the **23-sample** set (93M excluded), using the filtered/united objects from
`26-methylkit-object.Rmd`. Implements:

- Per-CpG **DML** via logistic regression with **overdispersion correction**
  (`overdispersion = "MN"`, `test = "Chisq"`) and **site as covariate** to
  mitigate pseudoreplication / site–exposure confounding.
- Tiled-window **DMRs** (1 kb) with the same model.
- Headline thresholds **q < 0.01 and |meth diff| >= 55%** (DML) / **>= 25%** (DMR),
  plus a 25%/50% effect-size spectrum.
- **BH-FDR** reported as headline alongside methylKit's default **SLIM**.
- **Per-site concordance** flag (same direction in all 3 sites of a group).
- 5x sensitivity pass.

Output dir: `../output/30-dml-dmr-methylkit/`

```{bash make-output-dir}
mkdir -p ../output/30-dml-dmr-methylkit
```

# Load filtered objects + metadata

```{r load}
meth_23_10x <- readRDS("../output/26-methylkit-object/meth_23_10x.rds")
filt_23_10x <- readRDS("../output/26-methylkit-object/filt_23_10x.rds")
filt_23_5x  <- readRDS("../output/26-methylkit-object/filt_23_5x.rds")
meta        <- read.csv("../output/26-methylkit-object/sample_metadata.csv")

# Align metadata to the 23-sample object order, drop 93M
meta_23 <- meta[match(meth_23_10x@sample.ids, meta$sample), ]
covariates <- data.frame(site = factor(meta_23$site))
covariates
```

# 2.2 Per-CpG DML — overdispersion + site covariate

```{r dml}
myDiff_cov <- calculateDiffMeth(
  meth_23_10x,
  covariates     = covariates,
  overdispersion = "MN",
  test           = "Chisq",
  mc.cores       = 24
)
saveRDS(myDiff_cov, "../output/30-dml-dmr-methylkit/myDiff_dml_10x.rds")
```

```{r dml-no-covariate}
# Comparison model WITHOUT site covariate (overdispersion only)
myDiff_nocov <- calculateDiffMeth(
  meth_23_10x,
  overdispersion = "MN",
  test           = "Chisq",
  mc.cores       = 24
)
saveRDS(myDiff_nocov, "../output/30-dml-dmr-methylkit/myDiff_dml_10x_nocov.rds")
```

# BH-FDR alongside SLIM

methylKit's `qvalue` column is SLIM; recompute BH from the raw p-values and use
BH as the headline.

```{r bh-fdr}
add_bh <- function(md) {
  d <- getData(md)
  d$qvalue_BH <- p.adjust(d$pvalue, method = "BH")
  d
}
dml_tab <- add_bh(myDiff_cov)
write.csv(dml_tab, "../output/30-dml-dmr-methylkit/dml_all_with_BH.csv", row.names = FALSE)
head(dml_tab)
```

# Effect-size spectrum + hyper/hypo (headline 55%)

```{r dml-thresholds}
# methylKit SLIM-based extraction at several effect sizes
dml_25      <- getMethylDiff(myDiff_cov, difference = 25, qvalue = 0.01)
dml_50      <- getMethylDiff(myDiff_cov, difference = 50, qvalue = 0.01)
dml_55      <- getMethylDiff(myDiff_cov, difference = 55, qvalue = 0.01)   # headline
dml_55_hyper <- getMethylDiff(myDiff_cov, difference = 55, qvalue = 0.01, type = "hyper")
dml_55_hypo  <- getMethylDiff(myDiff_cov, difference = 55, qvalue = 0.01, type = "hypo")

counts <- data.frame(
  set   = c("q<0.01 & >=25%", "q<0.01 & >=50%", "q<0.01 & >=55%",
            ">=55% hyper", ">=55% hypo"),
  n_DML = c(nrow(dml_25), nrow(dml_50), nrow(dml_55),
            nrow(dml_55_hyper), nrow(dml_55_hypo))
)
write.csv(counts, "../output/30-dml-dmr-methylkit/dml_counts_by_threshold.csv", row.names = FALSE)

# Headline table (write .tab to mirror code/11 idiom for downstream bedGraph)
write.table(getData(dml_55), "../output/30-dml-dmr-methylkit/myDiff_1055p.tab")
counts
```

# BH-based headline set (transparency)

```{r dml-bh-set}
dml_bh_55 <- dml_tab |>
  filter(qvalue_BH < 0.01, abs(meth.diff) >= 55)
write.csv(dml_bh_55, "../output/30-dml-dmr-methylkit/dml_q01_55p_BH.csv", row.names = FALSE)
nrow(dml_bh_55)
```

# 2.3 Tiled-window DMRs (1 kb)

```{r dmr}
filt_23_10x <- readRDS("../output/26-methylkit-object/filt_23_10x.rds")

tiles <- tileMethylCounts(filt_23_10x, win.size = 1000, step.size = 1000, cov.bases = 5)
meth_tiles <- unite(tiles, min.per.group = 5L, destrand = FALSE)

myDiff_dmr <- calculateDiffMeth(
  meth_tiles,
  covariates     = covariates,
  overdispersion = "MN",
  test           = "Chisq",
  mc.cores       = 24
)
saveRDS(myDiff_dmr, "../output/30-dml-dmr-methylkit/myDiff_dmr_1kb.rds")

dmr_25 <- getMethylDiff(myDiff_dmr, difference = 25, qvalue = 0.01)
write.table(getData(dmr_25), "../output/30-dml-dmr-methylkit/dmr_25p_1kb.tab")
nrow(dmr_25)
```

# 2.3 (optional) dmrseq cross-check

```{r dmrseq}
if (requireNamespace("dmrseq", quietly = TRUE) &&
    requireNamespace("bsseq", quietly = TRUE)) {
  library(dmrseq); library(bsseq)

  # Build a BSseq object from the 23 cov files (methylation + coverage matrices).
  meta_23 <- read.csv("../output/26-methylkit-object/sample_metadata.csv") |>
    subset(sample != "93M")
  cov_files <- file.path("../output/25-meth-extract",
                         paste0(meta_23$sample, ".CpG_report.merged_CpG_evidence.cov"))
  bs <- read.bismark(files = cov_files, colData = DataFrame(row.names = meta_23$sample,
                                                            PAH = factor(meta_23$pah)))
  bs <- bs[rowSums(getCoverage(bs) == 0) == 0, ]

  set.seed(1)  # determinism for permutation inference
  dmrs <- dmrseq(bs = bs, testCovariate = "PAH")
  saveRDS(dmrs, "../output/30-dml-dmr-methylkit/dmrseq_dmrs.rds")
  as.data.frame(dmrs) |> head()
}
```

# 2.5 Per-site concordance flag

A candidate DML should change in the **same direction in all 3 sites** of the
high group relative to the low baseline (guards against single-site/individual
drivers).

```{r per-site-concordance}
# Per-site mean methylation at the headline DML positions
percMeth <- percMethylation(meth_23_10x)               # CpG x sample matrix
site_vec <- meta_23$site
site_means <- sapply(unique(site_vec), function(s)
  rowMeans(percMeth[, site_vec == s, drop = FALSE], na.rm = TRUE))

low_sites  <- c("HC","AP","BS"); high_sites <- c("EB","SA","SC")
low_mean   <- rowMeans(site_means[, low_sites,  drop = FALSE], na.rm = TRUE)

# For each high site, sign of (high site - low baseline)
dir_by_site <- sapply(high_sites, function(s) sign(site_means[, s] - low_mean))
concordant  <- abs(rowSums(dir_by_site)) == length(high_sites)

concordance <- data.frame(
  chr = meth_23_10x$chr, start = meth_23_10x$start, end = meth_23_10x$end,
  concordant_all3_high = concordant
)
write.csv(concordance, "../output/30-dml-dmr-methylkit/per_site_concordance.csv",
          row.names = FALSE)
table(concordance$concordant_all3_high)
```

# 5x sensitivity pass

```{r sensitivity-5x}
meth_23_5x <- readRDS("../output/26-methylkit-object/meth_23_5x.rds")
meta_5x <- read.csv("../output/26-methylkit-object/sample_metadata.csv")
meta_5x <- meta_5x[match(meth_23_5x@sample.ids, meta_5x$sample), ]

myDiff_5x <- calculateDiffMeth(meth_23_5x,
                               covariates = data.frame(site = factor(meta_5x$site)),
                               overdispersion = "MN", test = "Chisq", mc.cores = 24)
dml_55_5x <- getMethylDiff(myDiff_5x, difference = 55, qvalue = 0.01)
write.table(getData(dml_55_5x), "../output/30-dml-dmr-methylkit/dml_55p_5x.tab")

data.frame(threshold = c("10x", "5x"),
           n_DML_55p = c(nrow(dml_55), nrow(dml_55_5x)))
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
