---
title: "31- DML/DMR annotation (Goal 2)"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

------------------------------------------------------------------------

# Overview — Goal 2.4

Annotates the DMLs/DMRs from `30-dml-dmr-methylkit.Rmd` against the
`GCF_036588685.1` GFF (`22-genome-prep.Rmd`). Converts the methylKit `.tab`
output to bedGraph (mirroring `code/12`/`code/14`), runs `bedtools intersect`
to recover overlapping `gene`/`mRNA`/`exon`/`CDS` features + LOC gene IDs, and
`bedtools closest` for intergenic loci (nearest gene + distance). Records the
genomic context per locus and maps LOC IDs to NCBI gene descriptions.

- DML: `../output/30-dml-dmr-methylkit/myDiff_1055p.tab`
- DMR: `../output/30-dml-dmr-methylkit/dmr_25p_1kb.tab`
- GFF + per-feature GFFs: `../output/22-genome-prep/genomic.gff`,
  `../output/28-genomic-feature-distribution/features/`
- Output dir: `../output/31-dml-dmr-annotation/`

```{bash make-output-dir}
mkdir -p ../output/31-dml-dmr-annotation
```

# Convert DML/DMR .tab -> bedGraph

```{bash tab-to-bedgraph}
out="../output/31-dml-dmr-annotation"

# DML headline set
in_dml="../output/30-dml-dmr-methylkit/myDiff_1055p.tab"
echo "track type=bedGraph name=\"DML_1055p\" description=\"meth diff\"" > "$out/dml_1055p.bedgraph"
awk 'NR>1 {gsub(/"/,"",$2); print $2"\t"$3-1"\t"$4"\t"$8}' "$in_dml" >> "$out/dml_1055p.bedgraph"

# DMR set
in_dmr="../output/30-dml-dmr-methylkit/dmr_25p_1kb.tab"
echo "track type=bedGraph name=\"DMR_25p\" description=\"meth diff\"" > "$out/dmr_25p.bedgraph"
awk 'NR>1 {gsub(/"/,"",$2); print $2"\t"$3-1"\t"$4"\t"$8}' "$in_dmr" >> "$out/dmr_25p.bedgraph"

head "$out/dml_1055p.bedgraph" "$out/dmr_25p.bedgraph"
```

# bedtools intersect against per-feature GFFs

```{bash intersect}
out="../output/31-dml-dmr-annotation"
feat="../output/28-genomic-feature-distribution/features"

for region in dml_1055p dmr_25p; do
  # Strip the track header before intersecting
  grep -v "^track" "$out/${region}.bedgraph" | sort -k1,1 -k2,2n > "$out/${region}.bed"

  for f in gene mRNA exon CDS; do
    bedtools intersect -a "$out/${region}.bed" -b "$feat/${f}.gff" -wa -wb \
      > "$out/${region}_in_${f}.tsv"
    echo "${region} x ${f}: $(wc -l < "$out/${region}_in_${f}.tsv")"
  done
done
```

# Nearest gene for intergenic loci (bedtools closest)

```{bash closest}
out="../output/31-dml-dmr-annotation"
feat="../output/28-genomic-feature-distribution/features"

# Sort gene GFF for closest
awk '$1!~/^#/' "$feat/gene.gff" | sort -k1,1 -k4,4n > "$out/gene.sorted.gff"

for region in dml_1055p dmr_25p; do
  bedtools closest -a "$out/${region}.bed" -b "$out/gene.sorted.gff" -d \
    > "$out/${region}_closest_gene.tsv"
done
head "$out/dml_1055p_closest_gene.tsv"
```

# Recover LOC gene IDs + descriptions

```{r recover-gene-ids}
library(rtracklayer)
gff   <- import("../output/22-genome-prep/genomic.gff")
genes <- gff[gff$type == "gene"]

gene_meta <- tibble(
  gene_id     = genes$gene,
  locus       = genes$ID,
  chr         = as.character(seqnames(genes)),
  start       = start(genes),
  end         = end(genes),
  description = if ("description" %in% names(mcols(genes))) genes$description else NA_character_
)
write.csv(gene_meta, "../output/31-dml-dmr-annotation/gene_metadata.csv", row.names = FALSE)
head(gene_meta)
```

```{r annotate-closest}
# Parse closest output, pull LOC id + description, classify context
parse_closest <- function(path) {
  cl <- read.table(path, sep = "\t", quote = "", fill = TRUE)
  attr_col <- ncol(cl) - 1   # GFF attribute field
  dist_col <- ncol(cl)
  tibble(
    chr      = cl[[1]], start = cl[[2]], end = cl[[3]], meth.diff = cl[[4]],
    gene_loc = sub('.*gene=([^;]+).*', '\\1', cl[[attr_col]]),
    distance = cl[[dist_col]]
  ) |>
    mutate(context = ifelse(distance == 0, "genic", "intergenic"))
}

dml_anno <- parse_closest("../output/31-dml-dmr-annotation/dml_1055p_closest_gene.tsv")
dmr_anno <- parse_closest("../output/31-dml-dmr-annotation/dmr_25p_closest_gene.tsv")

write.csv(dml_anno, "../output/31-dml-dmr-annotation/dml_annotated.csv", row.names = FALSE)
write.csv(dmr_anno, "../output/31-dml-dmr-annotation/dmr_annotated.csv", row.names = FALSE)

table(dml_anno$context)
head(dml_anno)
```

# Genomic-context breakdown (gene/exon/intron/promoter/intergenic)

```{r context-breakdown}
# Reuse the feature BEDs from step 28 to tag each DML with finer context.
library(GenomicRanges)
bed2gr <- function(p) { d <- read.table(p); GRanges(d$V1, IRanges(d$V2+1, d$V3)) }

feat_dir <- "../output/28-genomic-feature-distribution/features"
feats <- c(exon = "exon.bed", intron = "intron.bed",
           promoter = "promoter_1kb.bed", intergenic = "intergenic.bed")

dml_gr <- GRanges(dml_anno$chr, IRanges(dml_anno$start+1, dml_anno$end))
ctx <- sapply(names(feats), function(f)
  overlapsAny(dml_gr, bed2gr(file.path(feat_dir, feats[[f]]))))
dml_context <- cbind(dml_anno, as.data.frame(ctx))
write.csv(dml_context, "../output/31-dml-dmr-annotation/dml_context_breakdown.csv",
          row.names = FALSE)
colSums(ctx)
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
