---
title: "32- Functional & feature enrichment (Goal 2)"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

------------------------------------------------------------------------

# Overview — Goal 2.6

Tests whether DMLs/DMRs (`31-dml-dmr-annotation.Rmd`) are functionally or
positionally enriched, using the proper backgrounds:

- **Feature enrichment:** DML over-representation in gene bodies/exons vs the
  genome-wide covered-CpG background from `28-genomic-feature-distribution.Rmd`
  (Fisher/χ²).
- **GO enrichment:** DML/DMR genes vs the background of *all annotated genes
  carrying covered CpGs* (the only genes that could be detected). NCBI GO may be
  sparse — supplement with orthology (eggNOG-mapper / InterProScan) as a gap.
- **Targeted xenobiotic check:** explicitly query CYP450 / GST / sulfotransferase
  / ABC-transporter genes (manuscript found none — confirm/extend).

Output dir: `../output/32-functional-enrichment/`

```{bash make-output-dir}
mkdir -p ../output/32-functional-enrichment
```

# Feature enrichment (gene-body vs covered-CpG background)

```{r feature-enrichment}
# Background: covered CpGs per feature (step 28). Foreground: DML per feature.
feat_bg <- read.csv("../output/28-genomic-feature-distribution/feature_methylation_summary.csv")
dml_ctx <- read.csv("../output/31-dml-dmr-annotation/dml_context_breakdown.csv")

total_bg  <- read.table("../output/28-genomic-feature-distribution/covered_cpg_10x.sorted.bed") |> nrow()
total_dml <- nrow(dml_ctx)

fisher_feature <- function(feature) {
  bg_in  <- feat_bg$n_cpg[feat_bg$feature == feature]
  dml_in <- sum(dml_ctx[[feature]], na.rm = TRUE)
  mat <- matrix(c(dml_in, total_dml - dml_in,
                  bg_in,  total_bg  - bg_in), nrow = 2)
  ft <- fisher.test(mat)
  tibble(feature = feature, dml_in = dml_in, bg_in = bg_in,
         odds_ratio = unname(ft$estimate), p_value = ft$p.value)
}

feat_enrich <- map_dfr(c("exon","intron","promoter","intergenic"), fisher_feature) |>
  mutate(p_adj = p.adjust(p_value, "BH"))
write.csv(feat_enrich, "../output/32-functional-enrichment/feature_enrichment.csv",
          row.names = FALSE)
feat_enrich
```

# Build GO background: genes with covered CpGs

```{r go-background}
library(GenomicRanges); library(rtracklayer); library(methylKit)

myobj_23 <- readRDS("../output/26-methylkit-object/myobj_23.rds")
covered <- lapply(seq_along(myobj_23), function(i) {
  d <- getData(myobj_23[[i]]); d <- d[d$coverage >= 10, ]
  GRanges(d$chr, IRanges(d$start, width = 1))
})
covered_gr <- reduce(do.call(c, covered))

gff   <- import("../output/22-genome-prep/genomic.gff")
genes <- gff[gff$type == "gene"]
bg_genes <- unique(genes$gene[overlapsAny(genes, covered_gr)])
writeLines(bg_genes, "../output/32-functional-enrichment/background_genes.txt")
length(bg_genes)
```

# GO / functional enrichment

NCBI GO terms are pulled from the GFF `Ontology_term`/`Dbxref` where present;
where sparse, attach orthology-based GO via eggNOG-mapper / InterProScan
(see bash chunk). Enrichment is tested with a hypergeometric/Fisher framework
against `background_genes.txt`.

```{r go-enrichment}
dml_anno <- read.csv("../output/31-dml-dmr-annotation/dml_annotated.csv")
dmr_anno <- read.csv("../output/31-dml-dmr-annotation/dmr_annotated.csv")
dm_genes <- unique(c(dml_anno$gene_loc, dmr_anno$gene_loc))
dm_genes <- dm_genes[dm_genes %in% bg_genes]
writeLines(dm_genes, "../output/32-functional-enrichment/dm_genes.txt")

# gene -> GO mapping (term2gene / term2name) parsed from GFF or eggNOG output
go_map_path <- "../output/32-functional-enrichment/gene2go.tsv"  # cols: gene  GO
if (file.exists(go_map_path)) {
  gene2go <- read.table(go_map_path, header = TRUE, sep = "\t")

  if (requireNamespace("clusterProfiler", quietly = TRUE)) {
    library(clusterProfiler)
    ego <- enricher(
      gene          = dm_genes,
      universe      = bg_genes,
      TERM2GENE     = gene2go[, c("GO", "gene")],
      pAdjustMethod = "BH",
      qvalueCutoff  = 0.05
    )
    write.csv(as.data.frame(ego), "../output/32-functional-enrichment/go_enrichment.csv",
              row.names = FALSE)
    head(as.data.frame(ego))
  }
} else {
  message("No gene2go.tsv found - generate GO mapping (NCBI/eggNOG) first.")
}
```

Orthology-based GO (run once; populates `gene2go.tsv`):

```{bash eggnog-orthology}
# Extract protein FASTA for annotated genes, then map to GO via eggNOG-mapper.
# emapper.py -i proteins.faa -o mtros --itype proteins --cpu 24 \
#   --output_dir ../output/32-functional-enrichment
# Downstream: parse mtros.emapper.annotations -> gene2go.tsv (gene <tab> GO)
echo "Run eggNOG-mapper / InterProScan here to build gene2go.tsv if NCBI GO is sparse."
```

# Targeted xenobiotic / biotransformation gene check

```{r xenobiotic-check}
gene_meta <- read.csv("../output/31-dml-dmr-annotation/gene_metadata.csv")

xeno_patterns <- c("cytochrome P450", "CYP", "glutathione S-transferase", "GST",
                   "sulfotransferase", "ABC transporter", "ATP-binding cassette")
pat <- paste(xeno_patterns, collapse = "|")

xeno_genes <- gene_meta |>
  filter(str_detect(coalesce(description, ""), regex(pat, ignore_case = TRUE)) |
         str_detect(coalesce(gene_id, ""),     regex("CYP|GST|SULT|ABC", ignore_case = TRUE)))

dml_anno <- read.csv("../output/31-dml-dmr-annotation/dml_annotated.csv")
dmr_anno <- read.csv("../output/31-dml-dmr-annotation/dmr_annotated.csv")
dm_genes <- unique(c(dml_anno$gene_loc, dmr_anno$gene_loc))

xeno_hit <- xeno_genes |> mutate(is_DM = gene_id %in% dm_genes | locus %in% dm_genes)
write.csv(xeno_hit, "../output/32-functional-enrichment/xenobiotic_gene_check.csv",
          row.names = FALSE)

# Headline result: how many biotransformation genes are differentially methylated?
cat("Xenobiotic/biotransformation genes overlapping DML/DMR:",
    sum(xeno_hit$is_DM), "of", nrow(xeno_hit), "annotated\n")
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
