---
title: "33- Biomarker candidate ranking (Goal 2)"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(methylKit)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

------------------------------------------------------------------------

# Overview — Goal 2.7-2.8

Promotes robust loci/regions to **biomarker candidates** by applying the plan's
multi-criteria filter, and runs the **site-as-replicate (n = 3 vs 3) sensitivity
analysis** — the statistically honest unit of replication for a site/exposure
effect. Produces a ranked shortlist for a targeted assay.

Biomarker criteria:

| Criterion | Operational definition |
|-----------|------------------------|
| Effect size | \|meth diff\| >= 55% (DML) / >= 25% (DMR), q < 0.01 |
| Consistency | same direction in all 3 sites; survives site-as-covariate model |
| Coverage robustness | >=10x in >=5 individuals/group; persists in 5x pass |
| Biological plausibility | maps to an annotated gene / reproducible region |
| Assayability | tractable for targeted bisulfite seq or MS-PCR |

Inputs from `30`–`32`. Output dir: `../output/33-biomarker-candidates/`

```{bash make-output-dir}
mkdir -p ../output/33-biomarker-candidates
```

# Site-as-replicate sensitivity (n = 3 low vs 3 high)

Collapse to site-level methylation means and test at the site level.

```{r site-level}
meth_23_10x <- readRDS("../output/26-methylkit-object/meth_23_10x.rds")
meta <- read.csv("../output/26-methylkit-object/sample_metadata.csv")
meta_23 <- meta[match(meth_23_10x@sample.ids, meta$sample), ]

perc <- percMethylation(meth_23_10x)             # CpG x individual
sites <- unique(meta_23$site)
site_mean <- sapply(sites, function(s) rowMeans(perc[, meta_23$site == s, drop = FALSE], na.rm = TRUE))

low_sites  <- c("HC","AP","BS"); high_sites <- c("EB","SA","SC")
site_test <- data.frame(
  chr = meth_23_10x$chr, start = meth_23_10x$start, end = meth_23_10x$end,
  low_mean  = rowMeans(site_mean[, low_sites]),
  high_mean = rowMeans(site_mean[, high_sites])
) |>
  mutate(meth.diff = high_mean - low_mean)

# Per-CpG t-test across the 3 vs 3 site means
site_test$p_site <- vapply(seq_len(nrow(perc)), function(i) {
  lo <- site_mean[i, low_sites]; hi <- site_mean[i, high_sites]
  tryCatch(t.test(hi, lo)$p.value, error = function(e) NA_real_)
}, numeric(1))
site_test$q_site <- p.adjust(site_test$p_site, "BH")

write.csv(site_test, "../output/33-biomarker-candidates/site_level_sensitivity.csv",
          row.names = FALSE)
sum(site_test$q_site < 0.05 & abs(site_test$meth.diff) >= 25, na.rm = TRUE)
```

# Assemble candidate evidence table

```{r assemble}
dml_bh   <- read.csv("../output/30-dml-dmr-methylkit/dml_all_with_BH.csv")
conc     <- read.csv("../output/30-dml-dmr-methylkit/per_site_concordance.csv")
dml_anno <- read.csv("../output/31-dml-dmr-annotation/dml_annotated.csv")
dml_5x   <- read.table("../output/30-dml-dmr-methylkit/dml_55p_5x.tab", header = TRUE)

key <- function(d) paste(d$chr, d$start, d$end, sep = "_")

cand <- dml_bh |>
  mutate(k = key(dml_bh)) |>
  left_join(conc |> mutate(k = key(conc)) |> select(k, concordant_all3_high), by = "k") |>
  left_join(dml_anno |> mutate(k = key(dml_anno)) |> select(k, gene_loc, context, distance), by = "k") |>
  left_join(site_test |> mutate(k = key(site_test)) |> select(k, q_site), by = "k") |>
  mutate(in_5x = k %in% key(dml_5x))
```

# Apply biomarker criteria + rank

```{r criteria}
biomarkers <- cand |>
  mutate(
    pass_effect      = qvalue_BH < 0.01 & abs(meth.diff) >= 55,
    pass_consistency = coalesce(concordant_all3_high, FALSE),
    pass_robustness  = coalesce(in_5x, FALSE),
    pass_plausible   = !is.na(gene_loc) & gene_loc != "." & distance == 0,
    pass_site_level  = !is.na(q_site) & q_site < 0.05,
    n_criteria       = pass_effect + pass_consistency + pass_robustness +
                       pass_plausible + pass_site_level
  ) |>
  filter(pass_effect) |>
  arrange(desc(n_criteria), desc(abs(meth.diff)), qvalue_BH)

write.csv(biomarkers, "../output/33-biomarker-candidates/biomarker_candidates_full.csv",
          row.names = FALSE)

shortlist <- biomarkers |>
  filter(pass_consistency, pass_robustness, pass_plausible) |>
  select(chr, start, end, meth.diff, qvalue_BH, gene_loc, context,
         concordant_all3_high, in_5x, q_site, n_criteria) |>
  head(50)
write.csv(shortlist, "../output/33-biomarker-candidates/biomarker_shortlist.csv",
          row.names = FALSE)
shortlist
```

# Targeted-assay BED for the shortlist

```{r assay-bed}
shortlist <- read.csv("../output/33-biomarker-candidates/biomarker_shortlist.csv")
assay_bed <- shortlist |>
  transmute(chr, start = start - 1, end, name = gene_loc, score = round(meth.diff))
write.table(assay_bed, "../output/33-biomarker-candidates/biomarker_shortlist.bed",
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

# Honest framing

```{r framing, eval=TRUE}
cat("With 3 sites/group, PAH exposure is confounded with site identity.\n",
    "Shortlisted loci are ASSOCIATED with high-PAH sites, not proven PAH-causal.\n",
    "Site-level (n=3 vs 3) significance and all-3-site concordance are the\n",
    "strongest available guards; confirm top candidates with a mixed-effects\n",
    "model (site as random effect) and an independent targeted assay.\n")
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
