---
title: "34- IGV & visualization prep"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

------------------------------------------------------------------------

# Overview

Final visualization step: produces per-sample methylation bedGraphs for IGV,
DML/DMR tracks, and an IGV session XML focused on the top biomarker candidates
(`33-biomarker-candidates.Rmd`). Genome + annotation come from
`22-genome-prep.Rmd`.

- Genome: `../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa`
- Annotation: `../output/22-genome-prep/genomic.gff`
- Tracks from: `../output/30-dml-dmr-methylkit/`, `../output/31-dml-dmr-annotation/`,
  `../output/33-biomarker-candidates/`
- Output dir: `../output/34-igv-visualization/`

```{bash make-output-dir}
mkdir -p ../output/34-igv-visualization
```

# Per-sample methylation bedGraphs (5x / 10x)

```{bash per-sample-bedgraphs}
cov_dir="../output/25-meth-extract"
out="../output/34-igv-visualization"

for f in ${cov_dir}/*.CpG_report.merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  awk -F $'\t' 'BEGIN{OFS=FS} {if ($5+$6 >= 5)  print $1,$2,$3,$4}' "${f}" > "${out}/${STEM}_5x.bedgraph"
  awk -F $'\t' 'BEGIN{OFS=FS} {if ($5+$6 >= 10) print $1,$2,$3,$4}' "${f}" > "${out}/${STEM}_10x.bedgraph"
done

ls -lh ${out}/*.bedgraph | head
```

# Copy DML/DMR/biomarker tracks

```{bash copy-tracks}
out="../output/34-igv-visualization"
cp ../output/31-dml-dmr-annotation/dml_1055p.bedgraph "${out}/" 2>/dev/null
cp ../output/31-dml-dmr-annotation/dmr_25p.bedgraph   "${out}/" 2>/dev/null
cp ../output/33-biomarker-candidates/biomarker_shortlist.bed "${out}/" 2>/dev/null
ls -lh ${out}
```

# Generate IGV session XML for top loci

```{r igv-session}
genome_url <- "https://gannet.fish.washington.edu/seashell/bu-github/project-mytilus-methylation/output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa"
gff_url    <- "https://gannet.fish.washington.edu/seashell/bu-github/project-mytilus-methylation/output/22-genome-prep/genomic.gff"

shortlist <- read.csv("../output/33-biomarker-candidates/biomarker_shortlist.csv")
top <- shortlist[1, ]
locus <- sprintf("%s:%d-%d", top$chr, max(1, top$start - 1000), top$end + 1000)

session <- sprintf('<?xml version="1.0" encoding="UTF-8" standalone="no"?>
<Session genome="%s" locus="%s" version="8">
  <Resources>
    <Resource path="dml_1055p.bedgraph" type="bedgraph"/>
    <Resource path="dmr_25p.bedgraph" type="bedgraph"/>
    <Resource path="biomarker_shortlist.bed" type="bed"/>
    <Resource path="%s" type="gff"/>
  </Resources>
</Session>', genome_url, locus, gff_url)

writeLines(session, "../output/34-igv-visualization/biomarker_session.xml")
locus
```

# igvR programmatic session (optional)

```{r igvr}
if (requireNamespace("igvR", quietly = TRUE)) {
  library(igvR)
  igv <- igvR()
  setBrowserWindowTitle(igv, "Mytilus PAH biomarkers")

  setCustomGenome(
    igv,
    genomeName    = "PNRI_Mtr1.1.1",
    fastaURL      = "../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa",
    indexURL      = "../output/22-genome-prep/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.fa.fai",
    annotationURL = "../output/22-genome-prep/genomic.gff"
  )

  shortlist <- read.csv("../output/33-biomarker-candidates/biomarker_shortlist.csv")
  top <- shortlist[1, ]
  setLocus(igv, sprintf("%s:%d-%d", top$chr, top$start - 1000, top$end + 1000))

  loadBedGraphTrack(igv, name = "DML 10/55p",
                    file = "../output/31-dml-dmr-annotation/dml_1055p.bedgraph",
                    color = "blue", trackHeight = 50)
}
```

# Summary figure: DML counts + direction

```{r summary-figure}
dml <- read.csv("../output/30-dml-dmr-methylkit/dml_all_with_BH.csv") |>
  filter(qvalue_BH < 0.01, abs(meth.diff) >= 55) |>
  mutate(direction = ifelse(meth.diff > 0, "hyper (high-PAH)", "hypo (high-PAH)"))

p <- ggplot(dml, aes(direction, fill = direction)) +
  geom_bar() +
  labs(x = NULL, y = "DML count (q<0.01, >=55%)",
       title = "Differentially methylated loci by direction") +
  theme_minimal() + theme(legend.position = "none")
ggsave("../output/34-igv-visualization/dml_direction_barplot.png", p, width = 5, height = 4)
```

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
