---
title: "35- DML gene-model figure (Table 2)"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 12,
  fig.align = "center"
)
```

------------------------------------------------------------------------

# Overview

Publication-quality, multi-panel figure of the **10 annotated gene models**
that harbor the 12 differentially methylated loci (DMLs) reported in `paper.md`
Table 2, with each DML denoted on its gene model. Companion to
`30-dml-dmr-methylkit.Rmd` and `31-dml-dmr-annotation.Rmd`.

- DML headline set: `../output/11-methylkit-klone/myDiff_1055p.tab`
  (methylKit, q<0.01, |meth.diff|>=55%; positions are 1-based CpG coordinates)
- Reference: NCBI **GCF_036588685.1** (PNRI_Mtr1.1.1.hap1)
- Annotation: downloaded NCBI GFF (this script)
- Output dir: `../output/35-dml-gene-model-figure/`

`rtracklayer`/`GenomicFeatures` are not installed in this environment, so the
GFF is parsed directly with `data.table`/`awk`.

```{bash make-output-dir}
mkdir -p ../output/35-dml-gene-model-figure
```

# Download the NCBI GFF annotation

Downloads `GCF_036588685.1_..._genomic.gff` from the NCBI FTP if not already
present, then carves out the lines for the 10 genes of interest so the rest of
the script stays fast and reproducible.

```{bash download-gff}
out="../output/35-dml-gene-model-figure"
gff="$out/genomic.gff"

if [ ! -s "$gff" ]; then
  url="https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/036/588/685/GCF_036588685.1_PNRI_Mtr1.1.1.hap1/GCF_036588685.1_PNRI_Mtr1.1.1.hap1_genomic.gff.gz"
  curl -sS -L -o "$gff.gz" "$url"
  gunzip -f "$gff.gz"
fi
ls -la "$gff"
```

```{bash subset-gff}
out="../output/35-dml-gene-model-figure"

awk -F'\t' '
BEGIN{ s="LOC134725187 LOC134690221 LOC134714536 LOC134695637 LOC134707074 LOC134720671 LOC134711256 LOC134685297 LOC134712818 LOC134724475";
       n=split(s,g," "); for(i=1;i<=n;i++) want[g[i]]=1 }
/^#/{next}
{ if(match($9,/gene=[^;]+/)){ gname=substr($9,RSTART+5,RLENGTH-5); if(gname in want) print } }
' "$out/genomic.gff" > "$out/gene_models_raw.gff"

echo "feature lines: $(wc -l < "$out/gene_models_raw.gff")"
cut -f3 "$out/gene_models_raw.gff" | sort | uniq -c
```

# Parse gene models

```{r parse-gff}
library(data.table)
library(dplyr)
library(ggplot2)
library(patchwork)

out_dir <- "../output/35-dml-gene-model-figure"
raw <- fread(file.path(out_dir, "gene_models_raw.gff"), sep = "\t", header = FALSE,
             quote = "", col.names = c("chr","src","type","start","end",
                                       "score","strand","frame","attr"))

get_attr <- function(attr, key) {
  out <- rep(NA_character_, length(attr))
  m   <- regexpr(paste0(key, "=[^;]+"), attr)
  hit <- m > 0
  out[hit] <- sub(paste0(key, "="), "", regmatches(attr, m))
  out
}

raw[, gene := get_attr(attr, "gene")]
raw[, ID := get_attr(attr, "ID")]
raw[, Parent := get_attr(attr, "Parent")]
raw[, product := get_attr(attr, "product")]
raw[, description := get_attr(attr, "description")]

genes <- raw[type == "gene",
             .(gene, chr, start, end, strand,
               gene_desc = description, gene_id_full = ID)]

mrna  <- raw[type == "mRNA",
             .(gene, tx = ID, chr, start, end, strand, product)]

# exon / CDS counts per transcript to help pick the primary isoform
exon_counts <- raw[type == "exon", .N, by = .(tx = Parent)]
setnames(exon_counts, "N", "n_exon")

mrna <- merge(mrna, exon_counts, by = "tx", all.x = TRUE)
mrna[, span := end - start]

# Primary transcript = longest genomic span, tie-break by most exons, then ID
setorder(mrna, gene, -span, -n_exon, tx)
primary <- mrna[, .SD[1], by = gene]
primary[]
```

```{r collect-features}
prim_tx <- primary$tx

exons <- raw[type == "exon" & Parent %in% prim_tx,
             .(gene, tx = Parent, start, end, strand)]
cds   <- raw[type == "CDS"  & Parent %in% prim_tx,
             .(gene, tx = Parent, start, end, strand)]

# attach gene to features via primary transcript map
tx2gene <- setNames(primary$gene, primary$tx)
exons[, gene := tx2gene[tx]]
cds[,   gene := tx2gene[tx]]

cat("exons per primary transcript:\n"); print(exons[, .N, by = gene])
```

# DML set (Table 2) and coordinate verification

The 12 DMLs of Table 2 are pulled from `myDiff_1055p.tab` by matching q-values,
then each is checked for containment within its assigned gene span.

```{r dml-table}
dml <- data.table(
  qvalue       = c(5.77e-14, 1.72e-21, 1.16e-11, 2.92e-13, 4.17e-14,
                   1.10e-13, 7.20e-13, 8.98e-18, 6.02e-19, 1.06e-12,
                   1.32e-13, 2.52e-16),
  chr          = c("NC_086379.1","NC_086379.1","NC_086373.1","NC_086376.1","NC_086373.1",
                   "NC_086374.1","NC_086378.1","NC_086375.1","NC_086375.1","NC_086381.1",
                   "NC_086375.1","NC_086379.1"),
  pos          = c(47961734, 64017640, 90626338, 5631156, 105503509,
                   34742951, 6633121, 31938079, 43581688, 21212100,
                   90876604, 1603830),
  assigned_gene= c("LOC134725187","LOC134725187","LOC134690221","LOC134714536","LOC134695637",
                   "LOC134707074","LOC134720671","LOC134711256","LOC134711256","LOC134685297",
                   "LOC134712818","LOC134724475"),
  meth_diff    = c(56.45, -63.6, 56.32, 59, -57.86,
                   58.27, -55.24, -60.69, 58.2, -56.07,
                   55.88, -60.82),
  status_paper = c("Both","Both","Hyper","Hyper","Hypo",
                   "Both","Hypo","Both","Both","Hypo",
                   "Hyper","Hypo")
)

# direction from the sign of meth.diff (hyper = +, hypo = -)
dml[, direction := factor(ifelse(meth_diff >= 0, "Hyper-methylated", "Hypo-methylated"),
                          levels = c("Hyper-methylated", "Hypo-methylated"))]

# verify each DML against its assigned gene span
gspan <- genes[, .(assigned_gene = gene, g_chr = chr, g_start = start,
                   g_end = end, g_strand = strand, gene_desc)]
dml <- merge(dml, gspan, by = "assigned_gene", all.x = TRUE, sort = FALSE)
dml[, in_gene_span := (chr == g_chr & pos >= g_start & pos <= g_end)]

dml[, .(assigned_gene, chr, pos, meth_diff, direction, status_paper,
        g_start, g_end, in_gene_span)][order(assigned_gene, pos)]
```

```{r verify-summary}
cat("DMLs inside their assigned gene span:", sum(dml$in_gene_span), "of", nrow(dml), "\n\n")
cat("DMLs NOT inside assigned gene span:\n")
print(dml[in_gene_span == FALSE, .(assigned_gene, chr, pos, meth_diff)])
```

**Note.** Ten of the twelve Table 2 loci fall cleanly inside their assigned gene
body. The two "second" loci of the bidirectional pairs do **not**:
`NC_086379.1:64017640` (assigned to LOC134725187, which is at ~47.96 Mb) and
`NC_086375.1:43581688` (assigned to LOC134711256, which is at ~31.94 Mb) are
intergenic in this RefSeq annotation, ~16 Mb and ~11 Mb from their assigned
genes respectively. Their nearest genes are *protein chibby homolog 1-like* /
uncharacterized LOC134725551 and *solute carrier family 40 member 1-like*. They
are therefore plotted only as a verification flag in the summary table, not on
the (distant) assigned gene models.

# Build the figure

```{r panel-builder}
# warm = hyper, cool = hypo
dml_cols <- c("Hyper-methylated" = "#D6604D", "Hypo-methylated" = "#2166AC")
feat_cols <- c("CDS" = "grey25", "UTR (exon)" = "grey75")

short_name <- function(desc) {
  desc <- sub(", transcript variant.*$", "", desc)
  desc
}

build_panel <- function(gid) {
  g  <- genes[gene == gid]
  tx <- primary[gene == gid]
  ex <- exons[gene == gid]
  cd <- cds[gene == gid]
  dd <- dml[assigned_gene == gid & in_gene_span == TRUE]

  # plot coordinates in kb (absolute genomic position)
  x0 <- tx$start / 1e3
  x1 <- tx$end   / 1e3

  ymax_dml <- 1.15
  # invisible layer carrying both DML levels so every panel yields an
  # identical colour guide (patchwork then collapses it to one legend)
  dummy <- data.frame(
    direction = factor(c("Hyper-methylated", "Hypo-methylated"),
                       levels = c("Hyper-methylated", "Hypo-methylated")),
    x = x0, y = 0)

  p <- ggplot() +
    # intron line spanning the primary transcript
    geom_segment(aes(x = x0, xend = x1, y = 0, yend = 0),
                 linewidth = 0.4, colour = "grey40") +
    geom_point(data = dummy, aes(x = x, y = y, colour = direction),
               alpha = 0, shape = 18, size = 3) +
    # exon boxes (UTR height)
    geom_rect(data = ex,
              aes(xmin = start/1e3, xmax = end/1e3, ymin = -0.22, ymax = 0.22,
                  fill = "UTR (exon)"), colour = NA) +
    # CDS boxes (full height, drawn over exons)
    geom_rect(data = cd,
              aes(xmin = start/1e3, xmax = end/1e3, ymin = -0.38, ymax = 0.38,
                  fill = "CDS"), colour = NA)

  # strand direction arrow above the model
  arr_y <- 0.62
  if (tx$strand == "+") {
    p <- p + annotate("segment", x = x0, xend = x1, y = arr_y, yend = arr_y,
                      colour = "grey55", linewidth = 0.3,
                      arrow = arrow(length = unit(0.10, "cm"), type = "closed"))
  } else {
    p <- p + annotate("segment", x = x1, xend = x0, y = arr_y, yend = arr_y,
                      colour = "grey55", linewidth = 0.3,
                      arrow = arrow(length = unit(0.10, "cm"), type = "closed"))
  }

  # DML lollipops
  if (nrow(dd) > 0) {
    p <- p +
      geom_segment(data = dd,
                   aes(x = pos/1e3, xend = pos/1e3, y = 0.38, yend = ymax_dml,
                       colour = direction), linewidth = 0.6, show.legend = FALSE) +
      geom_point(data = dd,
                 aes(x = pos/1e3, y = ymax_dml, colour = direction),
                 size = 3, shape = 18, show.legend = FALSE) +
      geom_text(data = dd,
                aes(x = pos/1e3, y = ymax_dml + 0.22,
                    label = sprintf("%+.1f%%", meth_diff), colour = direction),
                size = 2.7, fontface = "bold", show.legend = FALSE)
  }

  lab <- sprintf("%s  \u2014  %s  (%s strand)",
                 gid, short_name(g$gene_desc), tx$strand)
  span_kb <- (tx$end - tx$start) / 1e3

  p +
    scale_fill_manual(values = feat_cols, name = "Gene structure",
                      breaks = c("CDS","UTR (exon)")) +
    scale_colour_manual(values = dml_cols, name = "DML methylation",
                        drop = FALSE) +
    guides(colour = guide_legend(
      override.aes = list(alpha = 1, shape = 18, size = 3, linetype = 0))) +
    scale_y_continuous(limits = c(-0.7, ymax_dml + 0.45), expand = c(0,0)) +
    labs(title = lab,
         x = sprintf("%s position (kb)   \u2022   transcript span %.1f kb",
                     g$chr, span_kb)) +
    theme_minimal(base_size = 10) +
    theme(
      plot.title = element_text(size = 9.2, face = "bold"),
      axis.title.x = element_text(size = 7.5, colour = "grey30"),
      axis.text.x = element_text(size = 7),
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.3),
      legend.position = "bottom"
    )
}
```

```{r assemble, fig.width=10, fig.height=12}
# order panels by paper functional grouping
gene_order <- c("LOC134725187","LOC134690221","LOC134714536","LOC134695637",
                "LOC134707074","LOC134720671",
                "LOC134711256","LOC134685297",
                "LOC134712818","LOC134724475")

panels <- lapply(gene_order, build_panel)

fig <- wrap_plots(panels, ncol = 2) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Gene models harboring differentially methylated loci (DMLs) in Mytilus trossulus mantle",
    subtitle = "High- vs low-\u03a316 PAH sites \u00b7 NCBI GCF_036588685.1 \u00b7 exons (boxes), CDS (dark), introns (line); lollipops mark DMLs (\u0394 methylation %)",
    caption = "DML positions from methylKit (q<0.01, |\u0394meth|\u226555%). Each panel drawn to its own bp scale.",
    theme = theme(plot.title = element_text(face = "bold", size = 13),
                  plot.subtitle = element_text(size = 9, colour = "grey30"))
  ) &
  theme(legend.position = "bottom")

ggsave(file.path(out_dir, "dml_gene_models.png"), fig,
       width = 10, height = 12, dpi = 300, bg = "white")
ggsave(file.path(out_dir, "dml_gene_models.pdf"), fig,
       width = 10, height = 12, bg = "white")

fig
```

# Summary table

```{r summary-table}
# nearest-gene context for the two intergenic loci (computed from the GFF)
nearest_note <- c(
  "NC_086379.1:64017640" = "intergenic; nearest gene LOC134725550 (protein chibby homolog 1-like), ~3.7 kb",
  "NC_086375.1:43581688" = "intergenic; nearest gene LOC134711544 (solute carrier family 40 member 1-like), ~1.0 kb"
)

n_exon_map <- exons[, .N, by = gene]
setnames(n_exon_map, "N", "exon_count")

summary_tab <- dml |>
  merge(n_exon_map, by.x = "assigned_gene", by.y = "gene", all.x = TRUE, sort = FALSE) |>
  merge(primary[, .(assigned_gene = gene, primary_tx = tx)], by = "assigned_gene",
        all.x = TRUE, sort = FALSE)

summary_tab[, gene_span := paste0(g_start, "-", g_end)]
summary_tab[, locus := paste0(chr, ":", pos)]
summary_tab[, note := ifelse(in_gene_span, "DML within gene body",
                             nearest_note[locus])]

summary_out <- summary_tab[order(assigned_gene, pos),
  .(gene_id = assigned_gene, gene_description = gene_desc, chr,
    gene_span, strand = g_strand, primary_tx, exon_count,
    dml_position = pos, qvalue, meth_diff, direction,
    status_paper, in_gene_span, note)]

fwrite(summary_out, file.path(out_dir, "dml_gene_model_summary.tsv"), sep = "\t")
fwrite(summary_out, file.path(out_dir, "dml_gene_model_summary.csv"))
knitr::kable(summary_out)
```

# Session info

```{r session-info}
sessionInfo()
```
